---
title: "CosMx Spatial Transcriptomics Workflow"
author: "Replace with author list"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, message=FALSE, warning=FALSE}
source("../R/00_packages.R")
load_required_packages()
source("../R/cosmx_io.R")
source("../R/cosmx_metadata.R")
source("../R/cosmx_plotting.R")
source("../R/cosmx_niches.R")
source("../R/cosmx_de.R")

cfg <- read_config("../config/config.yml")
```

# Overview

This notebook provides a compact, publication-facing version of the CosMx/NanoString spatial transcriptomics workflow. Exploratory code, local paths, and one-off troubleshooting chunks have been moved into reusable scripts and functions.

# Load prepared Seurat object

```{r load-object}
obj <- read_cosmx_seurat(cfg$paths$prepared_seurat_rds)
Seurat::DefaultAssay(obj) <- cfg$seurat$assay
obj
```

# Metadata checks

```{r metadata-checks}
colnames(obj@meta.data)
summary(obj@meta.data[, intersect(c("HIV_status", "plaque_type", cfg$seurat$celltype_publication_col), colnames(obj@meta.data)), drop = FALSE])
```

# Spatial visualization

```{r spatial-plots, fig.width=8, fig.height=6}
slide_plots <- plot_all_images(obj, group_by = cfg$seurat$celltype_publication_col)
slide_plots[[1]]
```

# UMAP

```{r umap, fig.width=8, fig.height=6}
if ("umap" %in% names(obj@reductions)) {
  Seurat::DimPlot(obj, reduction = "umap", group.by = cfg$seurat$celltype_publication_col, label = TRUE) +
    ggplot2::theme_classic()
}
```

# Marker dot plot

```{r dotplot, fig.width=14, fig.height=6}
plot_marker_dotplot(obj, unlist(cfg$plotting$dotplot_genes), group_by = cfg$seurat$celltype_publication_col)
```

# Spatial niches

```{r niches, eval=FALSE}
niche_result <- compute_spatial_niches(
  obj,
  image = cfg$niches$fov,
  celltype_col = cfg$seurat$celltype_publication_col,
  kNN = cfg$niches$kNN,
  K = cfg$niches$K,
  max_cells = cfg$niches$max_cells,
  include_self = cfg$niches$include_self,
  seed = cfg$niches$seed
)

plot_spatial_niches(niche_result$niche_cells)
summarize_niche_composition(niche_result)
```

# Differential expression

```{r de, eval=FALSE}
de_results <- run_pairwise_de(
  obj,
  group_col = cfg$differential_expression$group_col,
  ident_1 = cfg$differential_expression$ident_1,
  ident_2 = cfg$differential_expression$ident_2,
  assay = cfg$seurat$assay
)

plot_volcano(de_results, genes_to_label = unlist(cfg$differential_expression$genes_to_label))
```

# Session information

```{r session-info}
sessionInfo()
```
