---
title: "CosMx Spatial Transcriptomics Analysis"
author: "C. Wanjalla"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    number_sections: false
    fig_caption: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

# Overview

This notebook provides a reproducible workflow for reviewing and analyzing
CosMx/NanoString spatial transcriptomics data stored as a Seurat object.
The workflow includes object loading, metadata review, sample annotation,
spatial visualization, molecule overlays, UMAP visualization, marker-gene
summary plots, spatial niche analysis, and differential expression.

# User-defined paths and settings

Edit the paths and variables in this section before running the notebook.
Avoid committing files that contain local paths or protected participant
identifiers to a public GitHub repository.

```{r user-settings}
# Input files
seurat_rds <- "~/Downloads/cosmx-transcriptomics-github-ready/CosMx_seurat_reviewer_ready2.rds"
giotto_rdata <- NA_character_  # optional; set to a valid .RData file if needed

# Output directory
output_dir <- "results"
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

# Core object settings
assay_use <- "Nanostring"
sample_id_col <- "Run_Tissue_name"
celltype_source_col <- "nb_clus"

# Spatial plotting settings
example_fov_index <- 1
example_fov_number <- 9

# Niche analysis settings
niche_fov <- NULL  # if NULL, first image is used
kNN <- 50
K <- 7
max_cells <- 8000
set.seed(1)
```

# Load required packages

```{r load-packages}
required_packages <- c(
  "Seurat", "SeuratObject", "dplyr", "ggplot2", "patchwork", "pheatmap",
  "Matrix", "ggrepel", "future", "viridis"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    "Please install missing packages before running this notebook: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(Seurat)
library(SeuratObject)
library(dplyr)
library(ggplot2)
library(patchwork)
library(pheatmap)
library(Matrix)
library(ggrepel)
library(future)
library(viridis)

future::plan("multisession", workers = 4)
options(stringsAsFactors = FALSE)
```

# Load CosMx Seurat object

```{r load-seurat-object}
if (!file.exists(seurat_rds)) {
  stop("Seurat object not found. Update `seurat_rds` in the user-settings chunk.")
}

cosmx_obj <- readRDS(seurat_rds)
DefaultAssay(cosmx_obj) <- assay_use
cosmx_obj
```

# Optional: load Giotto object

```{r load-giotto-object, eval=FALSE}
# This section is optional and should only be run if a Giotto object is needed.
# Keep it disabled for public/reviewer workflows unless Giotto-specific analyses are used.
if (!is.na(giotto_rdata) && file.exists(giotto_rdata)) {
  load(giotto_rdata)
  # Example checks, assuming the Giotto object is named `gem`:
  # show(gem)
  # head(gem@spatial_locs)
}
```

# Review object structure and metadata

```{r review-object-structure}
# Review metadata columns
colnames(cosmx_obj@meta.data)

# Review sample distribution
if (sample_id_col %in% colnames(cosmx_obj@meta.data)) {
  table(cosmx_obj@meta.data[[sample_id_col]], useNA = "ifany")
} else {
  stop("Sample ID column not found: ", sample_id_col)
}

# Review image/FOV names
image_names <- Images(cosmx_obj)
image_names
```
```{r}
standardize_metadata_column <- function(obj, target_col) {
  md_cols <- colnames(obj@meta.data)

  if (target_col %in% md_cols) {
    return(obj)
  }

  possible_cols <- c(
    paste0(target_col, ".x"),
    paste0(target_col, ".y"),
    paste0(target_col, "_x"),
    paste0(target_col, "_y")
  )

  matched_col <- possible_cols[possible_cols %in% md_cols][1]

  if (!is.na(matched_col)) {
    message("Using metadata column '", matched_col, "' as '", target_col, "'.")
    obj[[target_col]] <- obj[[matched_col]][, 1]
  } else {
    warning("Could not find metadata column: ", target_col)
  }

  obj
}

cosmx_obj <- standardize_metadata_column(cosmx_obj, "HIV_status")
cosmx_obj <- standardize_metadata_column(cosmx_obj, "plaque_type")
```

```{r}
# Normalize sample names
cosmx_obj$Run_Tissue_name_norm <- gsub(
  "_",
  ".",
  as.character(cosmx_obj$Run_Tissue_name)
)

# Define plaque-stage groups
late_samples <- c(
  "R6101.1657DLAD",
  "R6101.43PLAD",
  "R6101.5394PLAD",
  "R6101.5433LOM"
)

early_samples <- c(
  "R6119.S1.2225",
  "R6119.S2.4666",
  "R6119.S3.5160",
  "R6119.S4.5521"
)

# Assign plaque type
cosmx_obj$plaque_type <- ifelse(
  cosmx_obj$Run_Tissue_name_norm %in% early_samples,
  "Early",
  ifelse(
    cosmx_obj$Run_Tissue_name_norm %in% late_samples,
    "Late",
    NA
  )
)

cosmx_obj$plaque_type <- factor(
  cosmx_obj$plaque_type,
  levels = c("Early", "Late")
)

table(cosmx_obj$plaque_type, useNA = "ifany")
```


```{r review-assay-layers}
# Review assay layers and available dimensions
if (assay_use %in% names(cosmx_obj@assays)) {
  SeuratObject::Layers(cosmx_obj[[assay_use]])
}

# Show a small expression matrix preview using the first available layer
available_layers <- SeuratObject::Layers(cosmx_obj[[assay_use]])
layer_use <- if ("counts" %in% available_layers) "counts" else available_layers[1]
expr_preview <- GetAssayData(cosmx_obj, assay = assay_use, layer = layer_use)
expr_preview[1:min(6, nrow(expr_preview)), 1:min(6, ncol(expr_preview))]
```

# Add publication-ready metadata columns

Edit the mapping files below before publication. For public GitHub deposition,
consider using de-identified sample labels such as `Sample_01`, `Sample_02`, etc.

```{r annotate-metadata}
# Create publication cell-type label column.
# Replace this map with the final curated cell-type annotations.
celltype_map <- c(
  "a" = "Macrophage 2",
  "b" = "COL9A3+ cells",
  "c" = "B cells",
  "d" = "Fibroblast 2",
  "e" = "FN1+ cells",
  "f" = "TCF7+ Monocytes"
)

if (!celltype_source_col %in% colnames(cosmx_obj@meta.data)) {
  stop("Cell-type source column not found: ", celltype_source_col)
}

cosmx_obj$celltype_pub <- as.character(cosmx_obj@meta.data[[celltype_source_col]])
cosmx_obj$celltype_pub[cosmx_obj$celltype_pub %in% names(celltype_map)] <-
  celltype_map[cosmx_obj$celltype_pub[cosmx_obj$celltype_pub %in% names(celltype_map)]]
cosmx_obj$celltype_pub[is.na(cosmx_obj$celltype_pub) | cosmx_obj$celltype_pub == ""] <- "Unknown"
cosmx_obj$celltype_pub <- factor(cosmx_obj$celltype_pub)
Idents(cosmx_obj) <- "celltype_pub"

table(cosmx_obj$celltype_pub, useNA = "ifany")
```

```{r annotate-groups}
# De-identify or externalize this mapping before public GitHub deposition.
# Recommended: store sample annotations in data/sample_metadata.csv and read them here.

cosmx_obj$HIV_status <- factor(cosmx_obj$HIV_status, levels = c("PWoH", "PLWH"))
cosmx_obj$plaque_type <- factor(cosmx_obj$plaque_type, levels = c("Early", "Late"))

table(cosmx_obj$HIV_status, useNA = "ifany")
table(cosmx_obj$plaque_type, useNA = "ifany")
```

# Prepare assay layers

```{r join-and-normalize-layers}
# Join Seurat v5 layers when multiple per-image count layers are present.
cosmx_obj <- JoinLayers(cosmx_obj, assay = assay_use)

# Normalize if a normalized data layer is absent.
available_layers <- SeuratObject::Layers(cosmx_obj[[assay_use]])
if (!"data" %in% available_layers) {
  cosmx_obj <- NormalizeData(
    cosmx_obj,
    assay = assay_use,
    normalization.method = "LogNormalize",
    scale.factor = 1e4
  )
}

saveRDS(cosmx_obj, file.path(output_dir, "cosmx_seurat_prepared.rds"))
```

# Spatial visualization

## Plot all slide images

```{r plot-all-images, fig.height=7, fig.width=8}
plot_image <- function(obj, image_name, group_by = NULL) {
  ImageDimPlot(
    obj,
    fov = image_name,
    group.by = group_by,
    axes = TRUE,
    border.color = NA
  ) + ggtitle(image_name)
}

spatial_plots <- lapply(image_names, plot_image, obj = cosmx_obj, group_by = "celltype_pub")
spatial_plots[[1]]
```

```{r save-all-image-plots, eval=FALSE}
for (img in image_names) {
  p <- plot_image(cosmx_obj, img, group_by = "celltype_pub")
  ggsave(
    filename = file.path(output_dir, paste0("spatial_", img, ".png")),
    plot = p,
    width = 8,
    height = 7,
    dpi = 300
  )
}
```

## Helper for FOV-level zooms

```{r spatial-helper-functions}
get_zoom_limits <- function(obj, image_name, sample_id, fov_number, sample_col = sample_id_col) {
  cells_of_interest <- obj$id[
    obj$fov == fov_number & obj@meta.data[[sample_col]] == sample_id
  ]

  centroid_data <- obj@images[[image_name]]$centroids
  zoom_fov <- apply(
    centroid_data@coords[centroid_data@cells %in% cells_of_interest, , drop = FALSE],
    2,
    range,
    na.rm = TRUE
  )

  zoom_fov
}

plot_molecule_overlay <- function(
  obj,
  image_name,
  zoom_limits = NULL,
  molecules = c("CXCL12", "CXCR4", "ACKR3"),
  alpha = 0.2,
  mols_size = 0.8,
  nmols = 1e5
) {
  p <- ImageDimPlot(
    obj,
    fov = image_name,
    border.color = NA,
    alpha = alpha,
    molecules = molecules,
    mols.size = mols_size,
    nmols = nmols,
    axes = FALSE
  ) +
    theme_classic()

  if (!is.null(zoom_limits)) {
    p <- p + coord_cartesian(
      xlim = zoom_limits[, 2],
      ylim = zoom_limits[, 1],
      expand = FALSE
    )
  }

  p
}
```

```{r example-molecule-overlay}
example_image <- image_names[example_fov_index]
example_sample <- unique(cosmx_obj@meta.data[[sample_id_col]])[example_fov_index]

zoom_limits <- get_zoom_limits(
  obj = cosmx_obj,
  image_name = example_image,
  sample_id = example_sample,
  fov_number = example_fov_number,
  sample_col = sample_id_col
)

plot_molecule_overlay(
  obj = cosmx_obj,
  image_name = example_image,
  zoom_limits = zoom_limits,
  molecules = c("CXCL12", "CXCR4", "ACKR3", "CD163"),
  alpha = 0.1,
  mols_size = 1,
  nmols = 1e5
)
```

# UMAP and marker visualization

```{r umap-by-celltype}
DimPlot(
  cosmx_obj,
  group.by = "celltype_pub",
  label = TRUE,
  repel = TRUE
) +
  NoLegend() +
  ggtitle("Cell-type annotations")
```

```{r marker-dotplot}
marker_genes <- c(
  "TPM2", "ACTA2", "MYH11", "RGS5", "APOD", "MMP2", "CXCL14", "DCN",
  "LYZ", "HLA-DRA", "CD74", "C1QA", "C1QB", "C1QC", "FN1", "VCAN",
  "CD14", "TCF7", "ACKR3", "TNFSF14", "VWF", "PECAM1", "CCL21",
  "CXCL12", "CCL19", "SPP1", "JCHAIN", "IGHA1", "IGHG1", "IGKC"
)

marker_genes <- intersect(marker_genes, rownames(cosmx_obj))

DotPlot(cosmx_obj, features = marker_genes, group.by = "celltype_pub") +
  RotatedAxis() +
  ggtitle("Marker gene expression by cell type")
```

# Spatial niche analysis

```{r niche-helper-functions}
.pick_xy <- function(coords) {
  if (all(c("x", "y") %in% colnames(coords))) return(c("x", "y"))
  if (all(c("X", "Y") %in% colnames(coords))) return(c("X", "Y"))
  if (all(c("imagecol", "imagerow") %in% colnames(coords))) return(c("imagecol", "imagerow"))
  if (all(c("col", "row") %in% colnames(coords))) return(c("col", "row"))
  stop("Could not identify coordinate columns in GetTissueCoordinates().")
}

.find_cell_id_col <- function(coords, seurat_cells) {
  common <- intersect(
    colnames(coords),
    c("cell", "Cell", "cell_id", "cellID", "barcode", "Barcode", "spot", "Spot")
  )

  for (cc in common) {
    if (sum(as.character(coords[[cc]]) %in% seurat_cells) > 0) return(cc)
  }

  for (cc in colnames(coords)) {
    v <- coords[[cc]]
    if (!is.character(v) && !is.factor(v)) next
    if (sum(as.character(v) %in% seurat_cells) > 0) return(cc)
  }

  NULL
}

.knn_index <- function(X, k) {
  if (requireNamespace("FNN", quietly = TRUE)) {
    nn <- FNN::get.knn(X, k = k)
    return(list(idx = nn$nn.index, engine = "FNN"))
  }

  if (requireNamespace("RANN", quietly = TRUE)) {
    nn <- RANN::nn2(X, k = k + 1)
    return(list(idx = nn$nn.idx[, -1, drop = FALSE], engine = "RANN"))
  }

  stop("Install FNN or RANN before running niche analysis.")
}

compute_niches_for_fov <- function(
  obj,
  fov,
  celltype_col = "celltype_pub",
  kNN = 50,
  K = 7,
  max_cells = 8000,
  include_self = FALSE,
  seed = 1
) {
  md <- obj@meta.data
  if (!celltype_col %in% colnames(md)) {
    stop("Cell type column not found: ", celltype_col)
  }

  coords <- GetTissueCoordinates(obj, image = fov)
  if (is.null(coords) || nrow(coords) == 0) stop("No tissue coordinates found for: ", fov)

  seurat_cells <- colnames(obj)
  cell_id_col <- .find_cell_id_col(coords, seurat_cells)
  if (is.null(cell_id_col)) {
    stop("No coordinate column matched Seurat cell IDs for image: ", fov)
  }

  xy <- .pick_xy(coords)
  coords[[cell_id_col]] <- as.character(coords[[cell_id_col]])
  coords <- coords[coords[[cell_id_col]] %in% seurat_cells, , drop = FALSE]

  centroids <- aggregate(
    coords[, xy, drop = FALSE],
    by = list(cell = coords[[cell_id_col]]),
    FUN = mean,
    na.rm = TRUE
  )
  colnames(centroids) <- c("cell", "x", "y")
  rownames(centroids) <- centroids$cell

  if (nrow(centroids) < 100) {
    stop("Too few centroided cells for niche analysis: ", nrow(centroids))
  }

  set.seed(seed)
  if (nrow(centroids) > max_cells) {
    centroids <- centroids[sample(rownames(centroids), max_cells), , drop = FALSE]
  }

  cells_use <- rownames(centroids)
  ct <- as.character(md[cells_use, celltype_col])
  ct[is.na(ct) | ct == ""] <- "Unknown"
  ct_levels <- sort(unique(ct))
  if (length(ct_levels) < 2) stop("Need at least two cell types to compute niches.")

  X <- as.matrix(centroids[, c("x", "y"), drop = FALSE])
  kNN <- min(kNN, nrow(X) - 1)
  kNN <- max(kNN, 5)

  nn <- .knn_index(X, k = kNN)
  idx <- nn$idx

  comp_mat <- matrix(
    0,
    nrow = nrow(idx),
    ncol = length(ct_levels),
    dimnames = list(cells_use, ct_levels)
  )

  ct_vec <- setNames(ct, cells_use)
  for (i in seq_len(nrow(idx))) {
    nbr_cells <- cells_use[idx[i, ]]
    if (include_self) nbr_cells <- c(nbr_cells, cells_use[i])
    comp_mat[i, ] <- as.numeric(table(factor(ct_vec[nbr_cells], levels = ct_levels)))
  }

  comp_prop <- comp_mat / pmax(rowSums(comp_mat), 1)

  set.seed(seed)
  K <- min(K, nrow(comp_prop) - 1)
  K <- max(K, 2)
  km <- kmeans(comp_prop, centers = K, nstart = 25)

  niche <- paste0("Niche_", sprintf("%02d", km$cluster))
  names(niche) <- rownames(comp_prop)

  niche_cells <- data.frame(
    cell = cells_use,
    niche = niche[cells_use],
    celltype = ct_vec[cells_use],
    x = centroids[cells_use, "x"],
    y = centroids[cells_use, "y"],
    fov = fov,
    stringsAsFactors = FALSE
  )

  list(
    niche_cells = niche_cells,
    comp_prop = comp_prop,
    engine = nn$engine,
    params = list(fov = fov, kNN = kNN, K = K, max_cells = max_cells)
  )
}
```

```{r run-niche-analysis}
if (is.null(niche_fov)) niche_fov <- image_names[1]

niche_result <- compute_niches_for_fov(
  obj = cosmx_obj,
  fov = niche_fov,
  celltype_col = "celltype_pub",
  kNN = kNN,
  K = K,
  max_cells = max_cells
)

niche_cells <- niche_result$niche_cells

table(niche_cells$niche)
head(niche_cells)
```

```{r plot-niches}
ggplot(niche_cells, aes(x = x, y = y, color = niche)) +
  geom_point(size = 0.7, alpha = 0.9) +
  coord_fixed() +
  theme_classic(base_size = 14) +
  labs(
    title = paste0("Spatial niches: ", niche_fov),
    subtitle = paste0("K = ", K, "; kNN = ", kNN, "; engine = ", niche_result$engine),
    x = "x coordinate",
    y = "y coordinate",
    color = "Niche"
  )
```

```{r niche-composition}
tab <- table(niche_cells$celltype, niche_cells$niche)
tab_prop <- prop.table(tab, margin = 2) * 100

pheatmap(
  tab_prop,
  scale = "none",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  main = "Cell-type composition per niche"
)
```

```{r niche-gene-expression}
genes_use <- c("CXCL12", "ACKR3", "CD163", "CD68", "CXCR4", "APOE")
genes_use <- intersect(genes_use, rownames(cosmx_obj))

expr <- GetAssayData(cosmx_obj, assay = assay_use, layer = "data")
cells_use <- intersect(colnames(expr), niche_cells$cell)

if (length(genes_use) > 0 && length(cells_use) > 0) {
  E <- expr[genes_use, cells_use, drop = FALSE]
  niche_by_cell <- niche_cells$niche[match(cells_use, niche_cells$cell)]
  niches <- sort(unique(niche_by_cell))

  avg_mat <- sapply(niches, function(nm) {
    cs <- cells_use[niche_by_cell == nm]
    Matrix::rowMeans(E[, cs, drop = FALSE])
  })

  z <- t(scale(t(as.matrix(avg_mat))))
  z[is.na(z)] <- 0

  pheatmap(
    z,
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    main = paste0("Niche marker expression: ", niche_fov)
  )
}
```

# Differential expression

This section is optional. For manuscript-level inference, consider using
sample-aware/pseudobulk models rather than treating cells as independent
biological replicates.

```{r differential-expression, eval=FALSE}
# Example only: replace ident.1 and ident.2 with final group labels.
Idents(cosmx_obj) <- "HIV_status"

de_results <- FindMarkers(
  cosmx_obj,
  ident.1 = "PLWH",
  ident.2 = "PWoH",
  assay = assay_use,
  slot = "data"
)

de_results$gene <- rownames(de_results)
de_results$p_val_adj <- p.adjust(de_results$p_val, method = "BH")

write.csv(de_results, file.path(output_dir, "de_results_PLWH_vs_PWoH.csv"), row.names = FALSE)
```

```{r volcano-plot, eval=FALSE}
label_genes <- c("CXCL12", "CXCR4", "ACKR3")
label_df <- de_results[de_results$gene %in% label_genes, , drop = FALSE]

ggplot(de_results, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_text_repel(data = label_df, aes(label = gene), max.overlaps = Inf) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  theme_classic(base_size = 14) +
  labs(
    title = "Differential expression",
    x = "Average log2 fold change",
    y = "-log10 adjusted p-value"
  )
```
```{r volcano-plot-with-significant-genes}

library(dplyr)
library(ggplot2)
library(ggrepel)
library(readr)

# Identify fold-change column used by your Seurat version
fc_col <- if ("avg_log2FC" %in% colnames(de_results)) {
  "avg_log2FC"
} else if ("avg_logFC" %in% colnames(de_results)) {
  "avg_logFC"
} else {
  stop("Could not find avg_log2FC or avg_logFC in de_results.")
}

# Make sure gene names are present
if (!"gene" %in% colnames(de_results)) {
  de_results$gene <- rownames(de_results)
}

# Add plotting and significance columns
de_results <- de_results %>%
  mutate(
    neg_log10_padj = -log10(pmax(p_val_adj, 1e-300)),
    significance = case_when(
      p_val_adj < 0.05 & .data[[fc_col]] >= 1  ~ "Higher in PLWH",
      p_val_adj < 0.05 & .data[[fc_col]] <= -1 ~ "Higher in PWoH",
      p_val_adj < 0.05                         ~ "Significant, modest FC",
      TRUE                                     ~ "Not significant"
    )
  )

# List significant genes
significant_genes <- de_results %>%
  filter(p_val_adj < 0.05) %>%
  arrange(p_val_adj) %>%
  select(gene, all_of(fc_col), p_val, p_val_adj, significance)

significant_genes

# Save significant genes table
dir.create("results", showWarnings = FALSE)

write_csv(
  significant_genes,
  "results/significant_genes_PLWH_vs_PWoH.csv"
)

# Label selected genes
label_genes <- c("CXCL12", "CXCR4", "ACKR3", "CD163", "ABCA1", "CD209")
label_df <- de_results %>%
  filter(gene %in% label_genes)

# Volcano plot
ggplot(
  de_results,
  aes(x = .data[[fc_col]], y = neg_log10_padj, color = significance)
) +
  geom_point(alpha = 0.6, size = 1.6) +
  geom_text_repel(
    data = label_df,
    aes(label = gene),
    max.overlaps = Inf,
    size = 4
  ) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed"
  ) +
  geom_vline(
    xintercept = c(-1, 1),
    linetype = "dashed"
  ) +
  scale_color_manual(
    values = c(
      "Higher in PLWH" = "red",
      "Higher in PWoH" = "blue",
      "Significant, modest FC" = "orange",
      "Not significant" = "gray70"
    )
  ) +
  theme_classic(base_size = 14) +
  labs(
    title = "Differential expression: PLWH vs PWoH",
    x = "Average log2 fold change",
    y = "-log10 adjusted p-value",
    color = "Gene category"
  )
```
```


# Session information

```{r session-info}
sessionInfo()
```
